---
title: "R: Tidy Data & Data Import 3"
author: "Sakol Suethanapornkul"
date: "`r format(Sys.time(), '%d %B %Y')`"
output: 
  beamer_presentation:
    theme: Sakol
    keep_tex: true
urlcolor: blue
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
```


## Preamble {.t}

So far, we have worked with various packages in `tidyverse`, from `dplyr`---through `ggplot2`---to `readr`. Through these packages, we have learned how to import into R, subset and manipulate, and then visualize data.

&nbsp;

Let's go through the exercise we saw at the end of the previous class once again to refresh our memory...

```{r eval = TRUE, warning = FALSE, message = FALSE}
library(tidyverse)
```



## Exercise {.t}

| Let's return to Income.csv. We will work through this exercise slowly.

|   [1] Import Income.csv and name the data frame as `income`.



## Exercise {.t}

| Let's return to Income.csv. We will work through this exercise slowly.

|   [1] Import Income.csv and name the data frame as `income`.

``` {r}
income <- read_csv(file = "./Data/Income.csv")
```



## Exercise {.t}

| Let's return to Income.csv. We will work through this exercise slowly.

|   [1] Import Income.csv and name the data frame as `income`.
|   [2] Create a new data frame consisting of country names and 
|       years since 1980.
|   [3] Select China, Japan, Malaysia, Singapore, and Thailand.



## Exercise {.t}

| Let's return to Income.csv. We will work through this exercise slowly.

|   [1] Import Income.csv and name the data frame as `income`.
|   [2] Create a new data frame consisting of country names and 
|       years since 1980.
|   [3] Select China, Japan, Malaysia, Singapore, and Thailand.

``` {r}
income_sm <- income %>% 
  select(!c(`1800`:`1979`)
         ) %>% 
  filter(country %in% c("China", "Japan", "Malaysia",
                        "Singapore", "Thailand")
         )
```

**NOTE**: Since our column names do not begin with letters, we need back ticks (\` \`).



## Exercise {.t}

| Let's return to Income.csv. We will work through this exercise slowly.

|   [1] Import Income.csv and name the data frame as `income`.
|   [2] Create a new data frame consisting of country names and 
|       years since 1980.
|   [3] Select China, Japan, Malaysia, Singapore, and Thailand.
|   [4] Pivot columns of years to a new variable "`year`" and values 
|       of GDP to "`gdp`"



## Exercise {.t}

| Let's return to Income.csv. We will work through this exercise slowly.

|   [1] Import Income.csv and name the data frame as `income`.
|   [2] Create a new data frame consisting of country names and 
|       years since 1980.
|   [3] Select China, Japan, Malaysia, Singapore, and Thailand.
|   [4] Pivot columns of years to a new variable "`year`" and values 
|       of GDP to "`gdp`"

``` {r}
income_ln <- income_sm %>% 
  pivot_longer(cols = !country, 
               names_to = "year", values_to = "gdp")
```



## Quick review {.t}

Recall that `pivot_longer` and `pivot_wider` are asymmetrical. Note how you select columns with these two functions.

1) `pivot_longer`:

``` {r}
income_sm %>% 
  pivot_longer(cols = !country, 
               names_to = "year", values_to = "gdp")
```

&nbsp;

2) `pivot_wider`:

``` {r}
income_ln %>% 
  pivot_wider(names_from = year, values_from = gdp)
```



## Part I: `bind_rows()` and `bind_cols()` {.t}

One important aspect we have not yet discussed is how to combine several data frames together!

&nbsp;

Before we do that, let's create two tibbles, one for each "class". We will have three columns: `id`, `class` and `height`. Can you remember how to create tibbles? 



## Part I: `bind_rows()` and `bind_cols()` {.t}

Here is our tibble for class A: 

``` {r eval = TRUE}
cl_a <- tibble(id     = 1:5,
               class  = rep("A", times = 5),
               height = c(173, 176, 177, 180, 182) 
               )
cl_a
```



## Part I: `bind_rows()` and `bind_cols()` {.t}

And this is for class B: 

``` {r eval = TRUE}
cl_b <- tibble(id     = 10:14,
               class  = rep("B", times = 5),
               height = c(167, 171, 172, 176, 178) 
               )
cl_b
```



## Part I: `bind_rows()` and `bind_cols()` {.t}

Our two tibbles have the exact same columns (i.e., same column names). Combining them is easy, with `bind_rows()`:

``` {r}
class_tot <-  bind_rows(cl_a, cl_b)
```

&nbsp;

Or we could "rewrite" our existing tibbles with:

``` {r}
cl_a <-  bind_rows(cl_a, cl_b)
```



## Part I: `bind_rows()` and `bind_cols()` {.t}

Suppose our class A has another column `gender` which represents the gender of students:

``` {r eval = TRUE}
cl_a <- cl_a %>% 
  mutate(gender = c("M", "F", "F", "F", "M") )

cl_a
```



## Part I: `bind_rows()` and `bind_cols()` {.t}

\textcolor{gured}{EXERCISE}: What will happen if we combine the first class (i.e., `cl_a`) with the second class (i.e., `cl_b`)?



## Part I: `bind_rows()` and `bind_cols()` {.t}

\textcolor{gured}{ANSWER}: An empty cell in the column `gender` will be given `NA`.

``` {r eval = TRUE}
cl_a <- bind_rows(cl_a, cl_b)

head(cl_a, n = 7)
```



## Part I: `bind_rows()` and `bind_cols()` {.t}

But it's very much possible that our original tibbles may not have an "identification." For example: 

Class A: 

``` {r eval = TRUE}
cl_a <- tibble(id     = 1:5,
               height = c(173, 176, 177, 180, 182) )
```

&nbsp;

Class B:

``` {r eval = TRUE}
cl_b <- tibble(id     = 10:14,
               height = c(167, 171, 172, 176, 178) )
```

\textcolor{gured}{EXERCISE}: How can we combine them such that we know which class each part of a new tibble come from?



## Part I: `bind_rows()` and `bind_cols()` {.t}

\textcolor{gured}{ANSWER}: We can name a new column with `.id`:

``` {r eval = TRUE}
cl <- bind_rows(.id = "class", "A" = cl_a, "B" = cl_b)

head(cl, n = 7)
```



## Part I: `bind_rows()` and `bind_cols()` {.t}

While `bind_rows()` binds tibbles by rows, `bind_cols()` does so by columns. 

&nbsp;

Let's look at an example.

``` {r eval = TRUE}
cl_a1 <- tibble(id     = 1:5,
                class  = rep("A", times = 5),
                height = c(173, 176, 177, 180, 182)
                )

cl_a2 <- tibble(gender = c("M", "F", "F", "F", "M"),
                grade  = c(3.3, 3.4, 3.5, 3.8, 3.9)
                )
```



## Part I: `bind_rows()` and `bind_cols()` {.t}

In `bind_cols()`, we provide as arguments tibble names:

&nbsp;

``` {r eval = TRUE}
bind_cols(cl_a1, cl_a2)
```



## Part I: `bind_rows()` and `bind_cols()` {.t}

But `bind_cols()` can behave unexpectedly. Let's say we have the following tibbles:

&nbsp;

``` {r eval = TRUE}
cl_a1 <- tibble(id     = 1:5,
                class  = rep("A", times = 5),
                height = c(173, 176, 177, 180, 182)
                )

cl_a2 <- tibble(id     = 1:5,
                gender = c("M", "F", "F", "F", "M"),
                grade  = c(3.3, 3.4, 3.5, 3.8, 3.9)
                )
```



## Part I: `bind_rows()` and `bind_cols()` {.t}

Go ahead and combine the two tibbles by columns:


``` {r}
bind_cols(cl_a1, cl_a2)
```


&nbsp;

Let's apply what we have learned from `bind_rows()`. Do following codes fix the problem?


``` {r}
bind_cols(.id = "id_new", cl_a1, cl_a2)

bind_cols(.id = "id_new", "1" = cl_a1, "2" = cl_a2)
```



## Part II: `*_join()` {.t}

It's time we talked about some other important functions that help us deal with **relational data** in a more sophisticated way.

&nbsp;

To start, we look back at the two tibbles:



## Part II: `*_join()` {.t}

``` {r eval = TRUE}
head(cl_a1, n = 3)

head(cl_a2, n = 3)
```



## Part II: `*_join()` {.t}

Even with this simple example, we can see one critical element of **relational data**: `id` links the two tibbles together. This element is referred to as a "key." 



## Part II: `*_join()` {.t}

`id` serves as a **primary key** that identifies unique observations in `cl_a1`. It also is a **foreign key** to `cl_a2` linking the two tibbles together.

&nbsp;

We can use `left_join()` to combine the two tibbles together:

``` {r}
left_join(x, y, 
          by = " ", 
          suffix = c(" ", " "))
```



## Part II: `*_join()` {.t}

So, to join the two tibbles together, we use:

``` {r eval = TRUE}
left_join(cl_a1, cl_a2, by = "id")
```



## Part II: `*_join()` {.t}

If you want to "rewrite" `cl_a1`, you can do that with:

``` {r eval = TRUE}
cl_a1 <- cl_a1 %>% 
  left_join(cl_a2, by = "id")

cl_a1
```



## Part II: `*_join()` {.t}

`left_join()` keeps **all observations in x** (which is `cl_a1` in our case). Let's say our other tibble has information from students no. 4 to 8.

``` {r eval = TRUE}
cl_a2 <- tibble(id     = 4:8,
                gender = c("M", "F", "F", "F", "M"),
                grade  = c(3.3, 3.4, 3.5, 3.8, 3.9)
                )
```

&nbsp;

\textcolor{gured}{EXERCISE}: What will happen when we run the following code?

``` {r}
left_join(cl_a1, cl_a2, by = "id")
```



## Part II: `*_join()` {.t}

The opposite of `left_join()` is `right_join()` which keeps **all observations in y**. 

&nbsp;

``` {r eval = TRUE, echo = FALSE}
cl_a1 <- tibble(id     = 1:5,
                class  = rep("A", times = 5),
                height = c(173, 176, 177, 180, 182)
                )

cl_a2 <- tibble(id     = 4:8,
                gender = c("M", "F", "F", "F", "M"),
                grade  = c(3.3, 3.4, 3.5, 3.8, 3.9)
                )
```


``` {r eval = TRUE}
right_join(cl_a1, cl_a2, by = "id")
```



## Part II: `*_join()` {.t}

There is another kind of mutating joins, which is `inner_join()`. It keeps only observations that both tibbles share:

``` {r eval = TRUE}
inner_join(cl_a1, cl_a2, by = "id")
```



## Part II: `*_join()` {.t}

The opposite of `inner_join()` is `full_join()` which keeps **all observations in x and y**.

``` {r eval = TRUE}
full_join(cl_a1, cl_a2, by = "id")
```



## Part II: `*_join()` {.t}

In many cases, you may have duplicate keys. For instance,

``` {r eval = TRUE}
x <- tribble(
  ~key, ~val_x,
     1, "x1",
     2, "x2",
     2, "x3",
     1, "x4"
  )

y <- tribble(
  ~key, ~val_y,
     1, "y1",
     2, "y2"
  )
```



## Part II: `*_join()` {.t}

Don't worry. When you join these tibbles (i.e., `x` and `y`) together, one is "expanded":

``` {r eval = TRUE}
left_join(x, y, by = "key")
```



## Q \& A {.t}

