---
title: "R: Data Transformation 2"
author: "Sakol Suethanapornkul"
date: "`r format(Sys.time(), '%d %B %Y')`"
output: 
  beamer_presentation:
    theme: Sakol
    keep_tex: true
urlcolor: blue
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)

```



## Quiz {.t}

For this quiz, we will use Pew Research Center's American Trend Panel Wave survey. The survey is downloadable from Pew's [website](https://www.pewresearch.org/politics/datasets/). We will read in a SPSS file and remove labels from the file:

``` {r}
library(haven)

#read spss file with read_sav()

wtp59 <- read_sav("ATPW59.sav") %>% 
  zap_label() %>% 
  zap_labels()
```



## Quiz {.t}

Just to make things easier, we will change column names to snake cases. We will make use of the package called ``janitor``. Install the package and copy and paste the following command to R:

``` {r}
library(janitor)

wtp59 <- wtp59 %>% 
  clean_names() #or janitor::clean_names()
```



## Quiz {.t}

| With the data frame in place, complete the following steps:

| [1] Create a new data frame with just the questions from the W59 
|     survey (which end with ``_w59``);

| [2] Drop questions about Youtube (which begin with ``yt``) from the 
|     data frame;

| [3] Move the last two columns on weight (``weight``) to third & 
|     fourth columns before ``device_type_w59``;

| [4] Drop two questions about a presidential candidate's quality 
|     (``demdealfght`` and ``repdealfght``); and

| [5] Check how many columns are left. What is the first column? 
|     Last?



## Recap {.t}

Let's remind ourselves again about vectors:

``` {r}
nums <- seq(from = 100, to = 200, by = 1.25)
length(nums)

is.numeric(nums)
is.vector(nums)

ord <- seq(from = 2, to = length(nums), by = 2)
nums[ord]

nums[nums >= mean(nums)]
nums >= mean(nums)

nums[[length(nums)]]

```



## Recap {.t}

We focused on tibbles (aka data frames):

- How to create them;
- How to select columns from tibbles; and
- How to relocate columns



## Recap {.t}

We focused on tibbles (aka data frames):

- How to create them;
- How to select columns from tibbles; and
- How to relocate columns

We focused on two functions: ``select()`` and ``relocate()``:

``` {r}
library(tidyverse)
data(diamonds)

dia <- select(diamonds, starts_with("c") )

dia <- relocate(dia, price, .before = color)
```



## Recap {.t}

\textcolor{gured}{EXERCISE}: How do the two functions differ with respect to an end product?

``` {r}
select(diamonds, where(is.numeric) )

relocate(diamonds, where(is.numeric) )
```



## Recap {.t}

\textcolor{gured}{ANSWER}: ``select()`` subsets your data, narrowing them down to only those columns you specify, while ``relocate()`` returns a tibble with the same number of columns (and rows) as the original one.



## Part I: ``filter()`` {.t}

Previously, we saw ``select()`` and ``relocate()`` which work with columns. In this part, we will look at functions that help us pick observations (i.e., *rows*) from tibbles.



## Part I: ``filter()`` {.t}

``filter()`` allows you subset and retain **rows** that meet your condition(s). 

We are going to work with a new data set (``mpg``) and would like to select cars with 6 or more/less cylinders  (in ``cyl``):

``` {r}
data(mpg)
help(mpg)

filter(mpg, cyl > 6)

filter(mpg, cyl < 6)
```



## Part I: ``filter()`` {.t}

``filter()`` works with comparison operations we saw previously:

``` {r}
filter(mpg, manufacturer  == "nissan")
filter(mpg, manufacturer  %in% c("nissan", "subaru"))
filter(mpg, !manufacturer %in% c("nissan", "subaru"))
#notice !manufacturer

filter(mpg, hwy >= 22)

filter(mpg, between(hwy, 20, 30)) 
#this is equivalent to:
filter(mpg, hwy > 20 & hwy < 30)
```



## Part I: ``filter()`` {.t}

And of course, you can combine more than one condition in ``filter()``:

``` {r}
filter(mpg, year == 2008,  cyl == 6)   #same as
filter(mpg, year == 2008 & cyl == 6)

filter(mpg, year == 1999 | cyl != 6)
```



## Part I: ``filter()`` {.t}

\textcolor{gured}{EXERCISE}: Find cars that...

[1] are manufactured by Toyota or Honda;

[2] are made by Toyota or Honda and have 4 cylinders;

[3] are SUV and can get 15 miles or more per gallon in city;

[4] are four-wheel drive SUV; and

[5] run 17 miles an hour in city and 20 miles an hour on highway.



## Part I: ``filter()`` {.t}

\textcolor{gured}{ANSWER}:

``` {r}
filter(mpg, 
       manufacturer %in% c("toyota", "honda"))

filter(mpg, 
       manufacturer %in% c("toyota","honda") & cyl>4)

filter(mpg, class == "suv" & hwy > 20)

filter(mpg, class == "suv" & drv == "4") 

filter(mpg, cty > 17 & hwy > 20)
```

**NOTE**: ``drv`` is a character vector.



## Part I: ``filter()`` {.t}

One thing we haven't discussed more thoroughly is ``NA``:

``` {r}
#NA is unknown
NA > 5
NA == NA

nums <- c(15, 20, 23, 24, NA, 26, NA, 31)

# NA is contagious
mean(nums)
sd(nums)
max(nums)
```

\textcolor{gured}{EXERCISE}: How do we check if elements inside a vector are ``NA``?



## Part I: ``filter()`` {.t}

\textcolor{gured}{ANSWER}: We use ``is.na()``.

``` {r}
nums <- c(15, 20, 23, 24, NA, 26, NA, 31)

is.na(nums)
```

In several functions, you can remove ``NA`` with ``na.rm = TRUE``

``` {r}
mean(nums, na.rm = TRUE)
sd(nums, na.rm = TRUE)
```



## Part I: ``filter()`` {.t}

``filter()`` only includes rows that are ``TRUE`` of your condition(s). If you want rows with ``NA``, you'll need to use ``is.na()``:

``` {r}
#read in Pew's data, if you haven't done so

filter(wtp59, yearahed_w59 == 1) 
filter(wtp59, yearahed_w59 == 1 | is.na(yearahed_w59)) 

# what's going on here?
filter(wtp59, yearahed_w59 == 1 & is.na(yearahed_w59))
```



## Part II: ``arrange()`` {.t}

Just as ``relocate()`` sorts columns, ``arrange()``does so with **rows**:

``` {r}
arrange(mpg, cty)
```

Use ``desc()`` inside ``arrange()`` to reorder in descending order:

``` {r}
arrange(mpg, desc(cty))
```



## Part II: ``arrange()`` {.t}

You can reorder rows in more than one columns. ``arrange()`` reorders rows one column at a time, using the next column to break ties:

``` {r}
arrange(mpg, desc(cty), desc(hwy))
```



## Part II: ``arrange()`` {.t}

Again, we can combine ``filter()`` and ``arrange()`` together:

``` {r}
mpg_small <- filter(mpg, 
                    !manufacturer %in% c("toyota", 
                                         "nissan")
                    )

mpg_small <- arrange(mpg_small, desc(year), desc(cty)
                     )
```



## Part III: ``mutate()`` {.t}

So far, we have learned how to select columns and rows that meet our condition(s). For instance, we may create a new tibble by selecting columns and rows that are:

``` {r}
mpg_small <- select(mpg, starts_with("m"), drv:class)
mpg_small <- filter(mpg_small, class != "suv")
```



## Part III: ``mutate()`` {.t}

But in a number of cases, we'd like to "do" something with our existing tibbles. This often involves creating new columns. We can easily do that with ``mutate()``:

``` {r}
#create a column that repeats "EPA" 
mutate(mpg_small, source = "EPA")

#or add this column back to the tibble
mpg_small <- mutate(mpg_small, source = "EPA")
```

\textcolor{gured}{EXERCISE}: Why did we end up with ``EPA`` being repeated > 170 times?



## Part III: ``mutate()`` {.t}

If you recall, arithmetic operation is vectorized in R:

``` {r eval = TRUE}
(num1 <- 1:10)
```

``` {r eval = TRUE}
num1 + c(2,4)
```

``` {r eval = TRUE}
num1 + 15
```



## Part III: ``mutate()`` {.t}

R expands a shorter vector to match the length of a longer one. This means that our ``source = "EPA"`` is expanded to match the length of ``mpg_small``

Also, if you recall:

``` {r eval = TRUE}
c(5, 6, 7, 8, 9) + c(10, 11, 12, 13, 14)
```



## Part III: ``mutate()`` {.t}

So, what do you think will happen here?

```{r}
mutate(mpg_small, dif = hwy - cty)

#or add to an existing tibble
mpg_small <- mutate(mpg_small, dif = hwy - cty)
```



## Part III: ``mutate()`` {.t}

You can reference this newly created variable right inside ``mutate()``:

```{r}
mutate(mpg_small, dif = hwy - cty,
                  dif_percent = dif / 100
       )
```



## Part III: ``mutate()`` {.t}

\textcolor{gured}{EXERCISE}: Create new variables that...

[1] subtract ``hwy`` from the maximum value of ``hwy``;

[2] subtract ``hwy`` from the average of ``hwy``

Then, spend some time thinking about each answer you've come up. How might your code work?



## Part III: ``mutate()`` {.t}

\textcolor{gured}{ANSWER}: 

``` {r}
mutate(mpg, max_dif  = max(hwy) - hwy)

mutate(mpg, mean_dif = mean(hwy) - hwy)
```

If you know that, for example, ``hwy`` contains some ``NA``, it is safe to:

``` {r}
mutate(mpg, mean_dif = max(hwy, na.rm = TRUE) - hwy)

mutate(mpg, mean_dif = mean(hwy, na.rm = TRUE) - hwy)
```



## Part III: ``mutate()`` {.t}

How does this work? Let's return to vectors:

``` {r eval = TRUE}
nums <- 11:20

max(nums) #max() gives the max value

max(nums) - nums
```



## Part III: ``mutate()`` {.t}

In a lot of cases, arithmetic operations can serve us well (with numeric vectors).

But in some cases, we may need to create a new column that is based on some condition of an existing column.



## Part III: ``mutate()`` {.t}

That can be done with ``if_else()``:

``` {r if_else}
gen    <- rep(c("M", "F"), times = 5)
scores <- 11:20

#dplyr::if_else(condition, true, false)

if_else(gen == "F", "female", "male")
if_else(scores > mean(scores), "above", "below")
```

**NOTE**: Use ``if_else()`` when there are just two options 



## Part III: ``mutate()`` {.t}

| \textcolor{gured}{EXERCISE}: Create new variables for ``mpg_small`` that...

|   [1] re-code ``year`` from 1999 to "90s" and from 2008 to "00s";

|   [2] classify ``cty`` into "above" and "below" based on ``mean(cty)``; 
|   and

|   [3] change only "suv" in ``class`` into "big car"



## Part III: ``mutate()`` {.t}

\textcolor{gured}{ANSWER}:

``` {r}
mutate(mpg, decade = if_else(year == 1999, 
                             "90s", "00s")
       )

mutate(mpg, cty_ave = if_else(cty > mean(cty), 
                              "above", "below")
       )

mutate(mpg, class = if_else(class == "suv", 
                            "big car", class)
       )
```



## Part III: ``mutate()`` {.t}

In contrast to ``if_else()``, ``case_when()`` can match several conditions:

``` {r}
nums <- 1:30

case_when(nums %% 10 == 0 ~ "ten",
          nums %% 5 == 0 ~ "five",
          nums %% 3 == 0 ~ "three",
          TRUE ~ as.character(nums)
          )
```

**NOTE**: ``case_when()`` moves from the specific to the general, so you need to put the most general on the last line. 



## Part III: ``mutate()`` {.t}

\textcolor{gured}{EXERCISE}: What is going to happen here?

``` {r}
case_when(nums %% 10 == 0 ~ "ten",
          nums %% 5 == 0 ~ "five",
          nums %% 3 == 0 ~ "three"
          )
```



## Part III: ``mutate()`` {.t}

You can apply ``case_when()`` and ``mutate()`` in some useful ways:

``` {r}
fruit <- c("banana", "apple", "mango")
food  <- c("hamburger", "pizza", "steak")

dish  <- c("hamburger", "banana", "steak", "rice", 
           "dumpling", "apple", "sushi", "hamburger", 
           "rice", "papaya", "orange")

case_when(dish %in% fruit ~ "fruits",
          dish %in% food ~ "foods",
          TRUE ~ "others")
```




## Part III: ``mutate()`` {.t}

| \textcolor{gured}{EXERCISE}: Let's go back to the ``mpg`` data set. Some cars in the data frame are American brands while others are foreign brands.  

|   [1] American brands: chevrolet, dodge, ford, jeep, lincoln;

|   [2] Japanese brands: honda, toyota, nissan;

|   [3] German brands: audi, volkswagen

There are other brands that are not in this list. **Create a new column for countries of these car brands**.



## Q & A