---
title: "R: Visualization 2"
author: "Sakol Suethanapornkul"
date: "`r format(Sys.time(), '%d %B %Y')`"
output: 
  beamer_presentation:
    theme: Sakol
    keep_tex: true
urlcolor: blue
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)

```



## Review {.t}

Let's begin by loading ``tidyverse`` which will also call ``ggplot2``:

``` {r eval = TRUE, warning = FALSE, message = FALSE}
library(tidyverse)
```



## Review {.t}

So far, we have practiced creating a few plots with ``ggplot2``.

For every plot we make, we need to begin by setting up a canvas. This translates to (1) \textcolor{gured}{declaring a data frame} and (2) \textcolor{gured}{mapping variables to aesthetic attributes}.



## Review {.t}

So far, we have practiced creating a few plots with ``ggplot2``.

For every plot we make, we need to begin by setting up a canvas. This translates to (1) \textcolor{gured}{declaring a data frame} and (2) \textcolor{gured}{mapping variables to aesthetic attributes}.

``` {r}
ggplot(data    = mpg, 
       mapping = aes(x = displ, y = cty)
       )
```

Or

``` {r}
mpg %>% 
  ggplot(mapping = aes(x = displ, y = cty)
         )
```



## Review {.t}

After that, we apply \textcolor{gured}{geometric objects} on the canvas. So far, we have seen three such objects, namely ``geom_point()``, ``geom_smooth()``, and ``geom_text()``.

``` {r}
mpg %>% 
  ggplot(mapping = aes(x = displ, y = cty)
         ) +
  geom_point(aes(color = drv)) +
  geom_smooth(method = "lm", se = FALSE
              )
```

Notice the arguments of ``geom_smooth()``, which are not part of ``aes()``.



## Review {.t}

Recall that any aesthetic attribute declared at the top level (i.e., inside ``ggplot()``) is passed down to all geoms.

``` {r}
mpg %>% 
  ggplot(mapping = aes(x = displ, y = cty) ) +
  geom_point(aes(color = drv)) +
  geom_smooth(method = "lm", se = FALSE)
```

``` {r}
mpg %>% 
  ggplot(mapping = aes(x = displ, y = cty, 
                       color = drv) ) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE)
```



## Data frame & ``ggplot()`` {.t}

This also means that any data frame declared at the top level is passed down to all geoms... 

&nbsp;

...and each geom can take a different data frame!



## Data frame & ``ggplot()`` {.t}

\textcolor{gured}{EXERCISE}: What will you get from running the following sets of code? 

``` {r}
a <- mpg %>% 
  filter(manufacturer %in% c("audi", "nissan", 
                             "toyota", "honda")
         )
```

``` {r}
b <- a %>% 
  group_by(manufacturer) %>% 
  summarize(m = mean(cty)
            )
```



## Data frame & ``ggplot()`` {.t}

Now, we can have two different data frames in one plot!

``` {r}
ggplot(data  = a, 
       aes(x = manufacturer) ) +
  geom_point(aes(y = cty), 
             shape = 5 ) +
  geom_point(data  = b, aes(y = m), 
             color = "red", 
             shape = 2 )
```



## ``ggplot2`` {.t}

In this lecture, we will look at other aspects of ``ggplot2`` focusing on statistical transformation and faceting.



## ``geom_bar()`` {.t}

Previously, we worked with ``geom_point()`` where each individual observation is plotted. But there's a different kind of plot, where data are summarized before being plotted. A very good example of this is a bar chart:

``` {r}
diamonds %>% 
  ggplot(aes(x = cut) ) +
  geom_bar()
```

&nbsp;

But pay attention to the y axis. Where does \textcolor{gured}{count} come from? It isn't in our ``aes()``!



## ``geom_bar()`` {.t}

Geometric objects rely on *stat* (which is short for statistical transformation) to compute values. Each geom has its default stat (e.g., run ``?geom_bar`` and ``?geom_point`` to check default stats). So, 

``` {r}
diamonds %>% 
  ggplot(aes(x = cut) ) +
  geom_bar(stat = "count") #default; often omitted
```

is equivalent to:

``` {r}
diamonds %>% 
  ggplot(aes(x = cut) ) +
  stat_count()
```



## ``geom_bar()`` {.t}

If we dig deeper into ``stat_count()``, we will see that it has a default geom, which is ``geom = "bar"``!

This shows that ``geom_bar()`` and ``stat_count()`` are pretty much the same thing!

&nbsp;

So why is this remotely interesting, you might ask?



## ``geom_bar()`` {.t}

Having this understanding means that we are not tied only to ``geom_bar()`` when we want to present frequency counts. With ``stat_count()``, we can use a different geom to present the counts!

``` {r}
diamonds %>% 
  ggplot(aes(x = cut) ) +
  stat_count(geom = "point")
```

``` {r}
diamonds %>% 
  ggplot(aes(x = cut) ) +
  stat_count(geom = "point", shape = 4)
```



## ``geom_bar()`` {.t}

Knowledge about *stat* can also come in handy. In some cases, we may want to mess with how frequency counts are to be presented.

&nbsp;

For example,

``` {r}
cnt <- diamonds %>% 
  count(cut)
```

\textcolor{gured}{EXERCISE}: What does this code do? What object does ``cnt`` refer to?



## ``geom_bar()`` {.t}

\textcolor{gured}{EXERCISE}: Will these two chunks of code work? Why or why not?

``` {r}
cnt %>% 
  ggplot(mapping = aes(x = cut)) + 
  geom_bar()
```

``` {r}
cnt %>% 
  ggplot(mapping = aes(x = cut, y = n)) + 
  geom_bar()
```



## ``geom_bar()`` {.t}

If you want to map the values of ``n`` to *y*, you will need to provide a different option for ``stat``. A default option is ``stat = "count"``, but now we need:  

``` {r}
cnt %>% 
  ggplot(mapping = aes(x = cut, y = n)) + 
  geom_bar(stat = "identity")
```

which is equivalent to:

``` {r}
cnt %>% 
  ggplot(mapping = aes(x = cut, y = n)) + 
  stat_identity(geom = "bar")
```



## ``geom_bar()`` {.t}

If you'd like to sort the x axis by frequencies (or counts), use ``reorder()``. This works with ``geom_bar()`` and ``geom_boxplot()``:

``` {r}
cnt %>% 
  ggplot() + 
  geom_bar(aes(x = reorder(cut, -n), 
               y = n),
               stat = "identity")
```



## ``geom_bar()`` {.t}

But in many cases, we'd like to present frequency counts by group. For example, in the ``diamonds`` data set, we may want to count the number of cuts by diamond colors. The following code doesn't give us what we want:

``` {r}
diamonds %>% 
  ggplot(aes(x = cut)) +
  geom_bar()
```



## ``geom_bar()`` {.t}

But in many cases, we'd like to present frequency counts by group. For example, in the ``diamonds`` data set, we may want to count the number of cuts by diamond colors. The following code doesn't give us what we want:

``` {r}
diamonds %>% 
  ggplot(aes(x = cut)) +
  geom_bar()
```

We can add another aesthetic attribute to the ggplot call:

``` {r}
diamonds %>% 
  ggplot(aes(x = cut, fill = color)) +
  geom_bar()
```



## ``geom_bar()`` {.t}

But we get different diamond colors stacked on top of each other. We can change this with ``position``:

``` {r}
diamonds %>% 
  ggplot(aes(x = cut, fill = color)) +
  geom_bar(position = "dodge") 
  #or geom_bar(position = position_dodge())
  #or stat_count(position = "dodge")
```

Run ``?geom_bar`` and look at ``position`` option.



## ``geom_bar()`` {.t}

Recall that in ``ggplot2``, graphics can be created by stacking one layer on top of another. This means we can, for instance, plot error bars on top of a bar chart!

&nbsp;

Let's start by summarizing our data:

``` {r}
diamonds %>% 
  group_by(cut, color) %>% 
  summarize(m  = mean(x), sd = sd(x),
            n  = n(),     se = sd/sqrt(n)
            )
```



## ``geom_bar()`` {.t}

And pipe this tibble into ggplot:

``` {r}
diamonds %>% 
  group_by(cut, color) %>% 
  summarize(m  = mean(x), sd = sd(x),
            n  = n(),     se = sd/sqrt(n)
            ) %>% 
  ggplot(aes(x = cut, y = m, fill = color) ) +
  geom_bar(stat = "identity", position = "dodge")
```



## ``geom_bar()`` {.t}

Then, we add ``geom_errorbar()`` to the call:

``` {r}
diamonds %>% 
  group_by(cut, color) %>% 
  summarize(m  = mean(x), sd = sd(x),
            n  = n(),     se = sd/sqrt(n)
            ) %>% 
  ggplot(aes(x = cut, y = m, fill = color) ) +
  geom_bar(stat = "identity", position = "dodge") +
  geom_errorbar(aes(ymin = m - se, 
                    ymax = m + se
                    ), 
                position = "dodge"
                )
```

See [this site](http://www.sthda.com/english/wiki/ggplot2-error-bars-quick-start-guide-r-software-and-data-visualization) for more examples.



## Faceting {.t}

Thus far, we have learned how to create a single plot using different geoms. But oftentimes, we may need to subset our data before plotting them. It's easy to do that with ``ggplot2``.



## Faceting {.t}

We use ``facet_wrap()`` to split a plot. We can choose how the plot is split (i.e., by rows or columns).

``` {r}
mpg %>% 
  ggplot(aes(x = displ, y = cty)) +
  geom_point() +
  facet_wrap(~ drv)
```

``` {r}
mpg %>% 
  ggplot(aes(x = displ, y = cty)) +
  geom_point() +
  facet_wrap(~ drv, nrow = 3)
```



## Faceting {.t}

Another option is to use ``facet_grid()`` to split a plot by rows and columns. This can be done with a formula ``a ~ b``:

``` {r}
mpg %>% 
  ggplot(aes(x = displ, y = cty)) +
  geom_point() +
  facet_grid(year ~ drv)
```



## Faceting {.t}

You can adjust positions of x and y axes with ``scales``, which works with both ``facet_grid()`` and ``facet_wrap()``:

``` {r}
mpg %>% 
  ggplot(aes(x = displ, y = cty)) +
  geom_point() +
  facet_grid(year ~ drv,
             scales = "free_x")
```

The four options are: ``fixed``, ``free``, ``free_x``, ``free_y``



## To facet or not to facet {.t}

Faceting has its pros and cons. If there is not much overlap, it may be better to use aesthetic attributes instead of faceting:

``` {r}
mpg %>% 
  ggplot(aes(x = displ, y = cty, color = drv)) +
  geom_point(position = "jitter") 
  # or geom_jitter(width = ..., height = ...)
```

``` {r}
mpg %>% 
  ggplot(aes(x = displ, y = cty)) +
  geom_point(position = "jitter") +
  facet_wrap(~ drv)
```



## Themes and apperances {.t}

``ggplot2`` offers almost unlimited options when it comes to appearances. Let's start with labels (see [this site](http://www.cookbook-r.com/Graphs/Legends_(ggplot2)/) for more information):

``` {r}
mpg %>% 
  ggplot(aes(x = displ, y = cty, color = drv)) +
  geom_point(position = "jitter") +
  labs(x = "Engine size", y = "Miles per gallon", 
       color = "Car type")
```

Or

``` {r}
mpg %>% 
  ggplot(aes(x = displ, y = cty, color = drv)) +
  geom_point(position = "jitter") +
  labs(x = NULL, y = NULL, color = NULL)
```



## Themes and apperances {.t}

You can also adjust color palette. For instance, ColorBrewer offers great options for colors (see [this site](https://rdrr.io/cran/RColorBrewer/man/ColorBrewer.html)):

``` {r}
mpg %>% 
  ggplot(aes(x = displ, y = cty, color = drv)) +
  geom_point(position = "jitter") +
  labs(x = NULL, y = NULL, color = NULL) +
  scale_color_brewer(palette = "Set1")
```



## Themes and apperances {.t}

We can also set overall theme of our graphics with ``theme_`` such as:

``` {r}
mpg %>% 
  ggplot(aes(x = displ, y = cty, color = drv)) +
  geom_point(position = "jitter") +
  labs(x = NULL, y = NULL, color = NULL) +
  scale_color_brewer(palette = "Set1") +
  theme_bw()
```

Try: ``theme_dark()``, ``theme_light()``, or ``theme_classic()``



## Themes and apperances {.t}

And finally, we can customize all the tiny little details (non-data) in our plot with ``theme()``:

``` {r}
mpg %>% 
  ggplot(aes(x = displ, y = cty, color = drv)) +
  geom_point(position = "jitter") +
  labs(x = NULL, y = NULL, color = NULL) +
  scale_color_brewer(palette = "Set1") +
  theme_bw() +
  theme(legend.position = "top",
        legend.title = element_text(size = 9),
        legend.text = element_text(size = 9)
        )
```

Check out [this site](https://ggplot2.tidyverse.org/reference/theme.html).



## Saving graphics {.t}

We can save our plots with ``ggsave()``. By default, the most recent plot is saved:

``` {r}
ggsave("file_name.png", 
       dpi = 300, 
       width = ..., 
       height = ..., 
       units = "in") # or "cm" / "mm"
```



## Show & Tell {.t}

